Title of Paper

### 2017-02-17 paper.tex
pass 1

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform for single-instruction, multiple-data vectorized OpenCL chemical source term evaluation was developed and validated against Cantera for a wide range of chemical kinetic models.
Speedups of \SI{1.88}{2.42}{$\times$} were observed over a baseline single-instruction, multiple-thread parallelized code showing a clear performance benefit to use of a row-major data layout.
Larger vector widths were found to produce speedups up to \SI{1.8}{$\times$} for smaller models, but had similar performance to smaller vector widths as the model size increased.
Finally, the parallel scaling efficiency was found to be slightly better for larger models, and future extensions of this work were discussed.
\end{abstract}

% (Provide 2-4 keywords describing your research. Only abbreviations firmly
% established in the field may be used. These keywords will be used for
% sessioning/indexing purposes.)
\begin{keyword}
    Chemical Kinetics\sep Single Instruction, Multiple Data\sep Vector Processing
\end{keyword}

@@

Single Instruction, Multiple-Data (SIMD) processing, also known as vector-processing, is an important parallel computing paradigm used increasingly in scientific computing.
While central processing unit (CPU) clock speeds have increased rapidly over the past few decades---the phenomena known as Moore's Law---power consumption and head dissipation issues have caused this trend to slow.
Traditional multi-core parallelism is often used to increase CPU performance, however SIMD processors---and the related Single Instruction, Multiple Thread (SIMT) processors, e.g. graphics processing units (GPUs)---have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems (e.g. CPU, GPU, etc.).

@@

A SIMD instruction utilizes specialized vector processing hardware to execute the same floating point operation (e.g. multiplication, division, etc.) on multiple pieces of data concurrently (Fig.~\ref{F:SIMD}).
Nearly all modern CPUs have vector processing units capable of running SIMD instructions, typically capable of handling 2--4 concurrent double precision operations---known as the vector width.
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (or MIC), have been developed that have tens of cores with very wide vector widths (e.g. 4--8 double precision operations); these very wide vector widths are also expected on forthcoming Intel CPUs (the Skylake Xeon and Cannon Lake architectures).

\subsection{Data ordering and vectorization patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$ there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all $N$ thermo-chemical states are adjacent.
In Fig.~\ref{F:mem} this would correspond to storing $[C_0]_0$ in index \num{0}, $[C_0]_1$ in index \num{1} and so on, with $N_1 = N$, and $N_2 = N_s$, the number of species in the model.
In the ``C'' (C row-major) format, the storage is flipped with $[C_0]_0$ in index \num{0}, $[C_1]_0$ in index \num{1} etc., with $N_1 = N_s$ and $N_2 = N$.
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state, has been implemented.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for different states will load sequential memory locations in Fig.~\ref{F:mem} in memory however the $k+1$'th species concentration will be $N$ memory locations away, increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state, thus SIMD-lanes loading species concentrations $k\dots k+V$ (for a vector width of $V$) will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_s$, with $N_s \ll N$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).
In Sec.~\ref{S:results} the effect of this ordering on the shallow-SIMD vectorization performance will be studied.

\subsection{Previous works and goals of this tudy}
Several recent works, e.g. \cite{CurtisGPU:2017} have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including the CPU, GPU and the MIC.
The shallow-SIMD and SIMT-vectorization models were investigated in OpenCL and compared to an OpenMP baseline code that was deep vectorized by simple compiler hints (a.k.a pragmas).
The shallow vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance while the GPU was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code, a \SI{94}{\percent} parallel efficiency.

This work will study the performance of chemical source term evaluation of automatically generated shallow-SIMD vectorized codes for a wide range chemical kinetic models on the CPU.
The performance will be compared to a baseline SIMT vectorized code to determine the effective SIMD speedup.
Finally future extensions to this work will be detailed.

@@

In addition, the temperature rate of change using the constant-pressure assumption\footnote{Note: in this context, the ``constant-pressure assumption'' refers to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure reactive-flow simulation.} is evaluated as:
\begin{equation}
\frac{\text{d}[T]}{\text{d}t} = -\frac{\sum_{k}^{N_S} H^{\degree}_k \frac{\text{d}[C_k]}{\text{d}t}}{\sum_{k}^{N_S} [C_k] c_{p_k}^{\degree}}
\end{equation}
although the code is equally capable of a evaluating the temperature rate using a constant-volume assumption, we omit this here for brevity.

\subsection{Code Generation}
Code generation is handled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which provides an interface to OpenCL allowing for unit testing and changes of program structure, e.g. data ordering, vectorization, threading patterns etc.

@@

In computing the net rate of progress of reaction $i$ from the forward and reverse ROP: $R_{i} = R_{i}^{\prime} - R_{i}^{\prime\prime}$, precision can easily be lost as the net rate of progress may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{P}_{\text{min},i} = 2^{-p_i} \le \left\lvert 1 - \frac{R_{i}^{\prime}}{R_{i}^{\prime\prime}} \right\rvert \le 2^{-q_i} = \text{P}_{\text{max},i}
\label{e:bounds}
\end{equation}
with $q_i \le p_i$.
To obtain an estimate of the a-priori precision bounds ($\text{P}_{\text{min},i}$ and $\text{P}_{\text{max},i}$), the forward and reverse ROP from Cantera are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, the Python Decimal class) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera, the bounds on the percent error attributable to precision loss, $\text{P}_{\text{err},i}$ can be estimated as:
\begin{equation}
\text{P}_{i,\text{min}} = 100 * \frac{\text{P}_{\text{min},i}}{\left\lvert\ R_{i,\text{CT}} - R_{i}\right\rvert} \le
\text{P}_{i} \le 100 * \frac{\text{P}_{\text{max},i}}{\left\lvert R_{i,\text{CT}} - R_{i} \right\rvert} = \text{P}_{i,\text{max}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera.

For a more direct comparison, a relative error norm of a quantity $X_{i,j}$ over all states $j$, and indices $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X_{i}} = \left\lVert \frac{\left\lvert X_{i,j,\text{CT}} - X_{i,j}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{i,j,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
and the lower precision bound for the net rate of progress error norm is set to $P_{\text{min}}^{*} = P_{i,\text{min}}$ for the $i, j$ value that results in the maximum error $E_{R_i}$.

@@

In Table~\ref{T:error}, we see the results of this code as compared to Cantera on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net rate of progress is \numrange{7}{8} orders of magnitude larger.
We see here that the minimum estimated loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining this jump.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera---as this loss of precision only occurs when the net ROP are many orders of magnitude smaller than the forward or reverse rates, implying the reaction is in near-equilibrium.

@@

Figure~\ref{F:Veccomp} investigates the effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from memory locations increasingly far apart and the performance of all three vector widths becomes similar.
Finally in Fig.~\ref{F:SIMDscale} the strong parallel scaling of the chemical source term evaluation is examined for all four models for a ``C''-ordered shallow vectorization with vector-width of 16:
The scaling efficiency for decreases with the number of cores used for all models; interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have similar efficiencies with isopentanol having the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared resources e.g., a L3 cache, which was not pursued in this work for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.

>>>>

%====================================================================
\begin{abstract} % not to exceed 200 words
A code generation platform for single-instruction, multiple-data-vectorized (SIMD) chemical source term evaluation was developed and validated against Cantera for a wide range of chemical kinetic models.
Speedups of up to \SIrange{1.88}{2.42}{$\times$} were observed over a baseline single-instruction, multiple-thread (SIMT) parallelized code.
Additionally, use of a row-major data layout was found to increase both the SIMD and SIMT performance by up to \SIrange{1.96}{2.38}{$\times$} over a column-major data layout.
Larger vector widths were found to produce speedups up to \SI{1.8}{$\times$} for smaller models, but had similar performance to smaller vector widths as the model size increased.
Finally, the parallel scaling efficiency was found to be slightly better for larger models, and future extensions of this work were discussed.
\end{abstract}

% (Provide 2-4 keywords describing your research. Only abbreviations firmly
% established in the field may be used. These keywords will be used for
% sessioning/indexing purposes.)
\begin{keyword}
    Chemical Kinetics\sep SIMD \sep Vector Processing
\end{keyword}

@@

Single Instruction, Multiple-Data (SIMD) processing, also known as vector-processing, is an important parallel computing paradigm used increasingly in scientific computing.
While central processing unit (CPU) clock speeds have increased rapidly over the past few decades---the phenomena known as Moore's Law---power consumption and heat dissipation issues have caused this trend to slow.
Traditional multi-core parallelism is often used to increase CPU performance, however SIMD processors---and the related Single Instruction, Multiple Thread (SIMT) processors, e.g. graphics processing units (GPUs)---have gained recognition due to their increased floating operation throughput.
The parallel programming standard OpenCL~\cite{stone2010opencl} has further enabled adoption of vector-processing based codes in scientific computing by providing a common application program interface (API) for execution on heterogeneous systems (e.g. CPU, GPU, etc.).

@@

A SIMD instruction utilizes specialized vector processing hardware to execute the same floating point operation (e.g. multiplication, division, etc.) on a several different data concurrently (Fig.~\ref{F:SIMD}); the number of concurrent operations is known as the vector-width.
Nearly all modern CPUs have vector processing units capable of running SIMD instructions, typically capable of handling 2--4 concurrent double precision operations.
Additionally, specialized hardware accelerators, e.g. Intel's Xeon Phi co-processor (or MIC), have been developed that have tens of cores with very wide vector widths (e.g. 4--8 double precision operations); these very wide vector widths are also expected on forthcoming Intel CPUs (the Skylake Xeon and Cannon Lake architectures).

\subsection{Data ordering and vectorization patterns}
\label{S:data}
When storing arrays for a chemical kinetic model, e.g. the concentration of species $k$ for the $j$-th of thermo-chemical state $[C_k]_j$ there are two data-ordering choices.
In the ``F'' (Fortran column-major) format, the concentrations of a single species $k$ over all $N$ thermo-chemical states are adjacent.
In Fig.~\ref{F:mem} this would correspond to storing $[C_0]_0$ in index \num{0}, $[C_0]_1$ in index \num{1} and so on, with $N_1 = N$, and $N_2 = N_s$, the number of species in the model.
In the ``C'' (C row-major) format, the storage order is transposed with $[C_0]_0$ in index \num{0}, $[C_1]_0$ in index \num{1} etc., and $N_1 = N_s$, $N_2 = N$.
This ordering---along with the device (CPU, GPU, etc.) in question---have a large effect on the performance of SIMD\slash SIMT-vectorized algorithms.

In a \textit{shallow}-SIMD vectorization, each SIMD-lane evaluates chemical source terms for a different thermo-chemical state.
If the data is stored in ``F''-order, the SIMD-lanes accessing the concentration of species $k$ for states $j, j+1\ldots$ will load sequential locations in memory; the $k+1$th species concentration will be $N$ memory locations away however, increasing the likelihood of cache-misses on the CPU.
In a \textit{deep}-SIMD vectorization, a vector-unit cooperates to evaluate chemical source terms for a single thermo-chemical state $j$, thus SIMD-lanes loading species concentrations $k, k+1\ldots$ will access sequential memory locations if the data is stored in ``C''-order.
Further, in the ``C''-ordering the furthest difference between any two concentrations in a single thermo-chemical state is at most $N_s$, with $N_s \ll N$ in most cases; this greatly improved data-locality increases the chances of a cache-hit on the CPU.
However, a deep vectorization may result in SIMD-\textit{waste}---similar to thread-divergence in GPU-processing---caused by different SIMD-lanes executing different instructions (e.g. from differing for-loop bounds).

Finally, in a SIMT-vectorization, threads execute the same instruction over multiple data in lock-step without use of specialized vectorization instructions.
This is the paradigm used in GPU-processing, a more through review of which can be found in our previous work~\cite{CurtisGPU}; one key-exception in this case is that the algorithm executed to evaluate the chemical source terms is identical for all thermo-chemical states, thus there is no thread-divergence for a shallow SIMT-vectorization. 
In Sec.~\ref{S:results} the effect of the data-ordering on the shallow-SIMD\slash SIMT vectorization performance will be studied.

\subsection{Previous works and goals of this study}
Several recent works, e.g. \cite{CurtisGPU:2017} have investigated SIMT-based chemical kinetic integration on GPUs, however SIMD-based chemical kinetics have been less studied in comparison.
\textcite{stone2016} implemented a linearly-implicit fourth-order stiff Rosenbrock solver in the OpenCL for various platforms including the CPU, GPU and the MIC.
The shallow-SIMD and SIMT-vectorization models were implemented in OpenCL and compared to an OpenMP baseline code that was deep SIMD-vectorized by simple compiler hints (a.k.a pragmas).
The shallow-SIMD vectorization improved the integrator performance over the OpenMP baseline by \SIrange{2.5}{2.8}{$\times$} on the CPU and \SIrange{4.7}{4.9}{$\times$} on the MIC, while the GPU performance was \SIrange{1.4}{1.6}{$\times$} slower than the OpenMP baseline due to thread-divergence concerns.
\textcite{kroshko2013efficient} implemented a shallow-SIMD vectorized third order stiff Rosenbrock integrator for atmospheric chemistry on a Cell Broadband Engine---a specially designed vector processor---finding a speedup of \SI{1.89}{$\times$} over a serial version of the same code, a \SI{94}{\percent} parallel efficiency.

This work will study the performance of chemical source term evaluation of automatically generated shallow-SIMD vectorized codes for a wide range chemical kinetic models on the CPU.
The performance will be compared to a baseline shallow-SIMT vectorized code to determine the effective SIMD speedup.
Finally future extensions to this work will be detailed.

@@

In addition, the temperature rate of change using the constant-pressure assumption\footnote{Note: in this context, the ``constant-pressure assumption'' refers to evaluation within a reaction sub-step in the operator splitting scheme, rather than a general constant-pressure reactive-flow simulation.} is evaluated on a molar basis as:
\begin{equation}
\frac{\text{d}[T]}{\text{d}t} = -\frac{\sum_{k}^{N_S} H^{\degree}_k \frac{\text{d}[C_k]}{\text{d}t}}{\sum_{k}^{N_S} [C_k] c_{p_k}^{\degree}}
\end{equation}
where $H^{\degree}_k$ and $c_{p_k}^{\degree}$ are the molar specific enthalpy and specific heat respectively.
Although the code is equally capable of a evaluating the temperature rate using a constant-volume assumption, we omit this here for brevity.

\subsection{Code Generation}
Code generation is enabled by the python package \texttt{loo.py}~\cite{kloeckner_loopy_2014}, which translates user specified psuedo-code and data to OpenCL, allowing for unit testing and easy changes of program structure, e.g. data ordering, vectorization, threading patterns etc.

@@

In computing the net ROP of reaction $i$ from the forward and reverse ROP: $R_{i} = R_{i}^{\prime} - R_{i}^{\prime\prime}$, precision can easily be lost as the net ROP may be---particularly near chemical equilibrium---many orders of magnitude smaller than the forward or reverse rates.
A base-ten bound on loss of precision~\cite{goldberg1991every} can be found as:
\begin{equation}
\text{L}_{i,\text{min}} = 2^{-p_i} \le \left\lvert 1 - \frac{R_{i}^{\prime}}{R_{i}^{\prime\prime}} \right\rvert \le 2^{-q_i} = \text{L}_{i,\text{max}}}
\label{e:bounds}
\end{equation}
with $q_i \le p_i$.
To obtain an estimate of the precision bounds, the forward and reverse ROP from Cantera are substituted into Eq.~\eqref{e:bounds}.
Additionally a higher-precision number format (in this case, the Python Decimal class) is used in this computation to maintain accuracy.
Finally, when determining the error in the net ROP between this code and Cantera, the bounds on the percent error attributable to precision loss, $\text{P}_{i,\text{err}}$ can be estimated as:
\begin{equation}
\text{P}_{i,\text{min}} = 100 * \frac{\text{L}_{i,\text{min}}}{\left\lvert\ R_{i,\text{CT}} - R_{i}\right\rvert} \le
\text{P}_{i,\text{err}} \le 100 * \frac{\text{L}_{i,\text{max}}}{\left\lvert R_{i,\text{CT}} - R_{i} \right\rvert} = \text{P}_{i,\text{max}}
\label{e:rel_bound}
\end{equation}
where the \text{CT} subscript indicates values from Cantera.

For a more direct comparison, a relative error norm of a quantity $X_{i,j}$ over all states $j$, and indices $i$ was computed using the $L^{\infty}$ norm:
\begin{equation}
 E_{X_{i}} = \left\lVert \frac{\left\lvert X_{i,j,\text{CT}} - X_{i,j}\right\rvert}{\num{e-10} + \num{e-6} * \left\lvert X_{i,j,\text{CT}} \right\rvert} \right\rVert_{i,j,\infty}
\label{e:rel_err}
\end{equation}
and the lower precision bound for the net ROP error norm is set to $\text{P}_{\text{min}}^{*} = \text{P}_{i,\text{min}}$ for the $i, j$ value that results in the maximum error $E_{R_i}$.

@@

In Table~\ref{T:error}, we see the results of this code as compared to Cantera on a library of partially stirred reaction conditions (PaSR) described in our previous works~\cite{CurtisGPU:2017,Niemeyer:2016aa}.
Very close agreement is observed for the forward and reverse ROP for all models, however the net ROP is \numrange{7}{8} orders of magnitude larger.
We see here that the minimum estimated loss of precision is \SIrange{4}{40}{$\times$} the error in net ROP, explaining this jump.
The species rate error norm is similar in magnitude to that of the net ROP, however the temperature rate error again increases as it depends directly on the species rates.
We note that the above discussion does not imply that computation of the net ROP will cause large error in chemical kinetic integration---either in this code or Cantera---as this loss of precision only occurs when the forward and reverse ROP are nearly equal, implying the reaction is in near-equilibrium.

@@

The effect of changing the vector-width supplied to a ``C''-ordered shallow-SIMD vectorization on a single CPU core is investigated in Fig.~\ref{F:Veccomp}.
Significant speedups are seen with larger vector-widths for the smaller models, however as the model size increases data must be loaded from increasingly distant memory locations and the performance of all three vector widths becomes similar.
Finally in Fig.~\ref{F:SIMDscale} the strong parallel scaling of the chemical source term evaluation is examined for all four models with a ``C''-ordered shallow SIMD-vectorization and vector-width of 16:
Interestingly the efficiency for the hydrogen model decreases fastest, while the larger models have similar efficiencies; the isopentanol model has the highest efficiency in all cases.
This suggests that evaluation of the chemical-source terms for a single thermo-chemical state for the smaller models may not saturate the processing power of the CPU, some efficiency may be lost due to context switching within OpenCL itself.
Additionally, OpenCL allows assigning of work to devices by shared memory resources e.g., a L3 cache, which was not pursued in this study for simplicity; this should be investigated in future work to determine the effect on scaling efficiency.
